First Order Alternation

### 2018-07-29 quantifiers/draft.tex
ok

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. As a shorthand, we write $\stamp{Q}{\leq n}$ (resp.
$\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid q \in Q, i
  \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in [n]}$). For
a formula $\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
variables and state predicates (and only those symbols) are replaced
by their time-stamped counterparts. For a sequence of input events
$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ (in
$\A$) is defined as:
\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{i}\]
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\T$}}\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
    \text{ occurs on the $n$-th level in $\T$}}$, we have $\I_{\T},
  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)})
  \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
  execution semantics of $\A$. This is the case because the children
  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
  $(n-1)$-th level of $\T$ form a cube from
  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs on the
  right-hand side. Then it must be the case that, for some node on the
  $(n-1)$-th level of $\G$, labeled with a configuration
  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}

Next, we give a logical characterization of acceptance, relative to a
given sequence of input events $\alpha \in \Sigma^*$. To this end, we
extend the path formula $\pathform{\alpha}$ with a characterization of
the set of final states of $\A$:

\[\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot\]

\begin{lemma}\label{lemma:acceptance}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
  every word $w \in \Sigma[X]^*$, the following are
  equivalent: \begin{compactenum}
  \item\label{it1:lemma:acceptance} there exists an interpretation
    $\I$ such that $\I,w_\Data \models \accform{w_\Sigma}$, 
    %
  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
  \end{compactenum}
\end{lemma}
\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
  that $\I,w_\Data \models \accform{w_\Sigma}$. By Lemma
  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
  that \begin{inparaenum}[(i)]
    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
      that
    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
      final configurations only. \end{inparaenum} First, assume that
  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
  m < n$. Then there exists a node on the $m$-th level, labeled with
  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
  children. By the definition of the execution semantics of $\A$, we
  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
  for $\nu[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
  $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
  q(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{m+1}$, thus
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
  contradicts the fact that $\I=\I_{\T}$ and that
  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
  \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$, by a
  similar reasoning as in the above case, we obtain that
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.

  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
  of $\A$ over $w$. We prove that $\I_\T,w_\Data \models
  \accform{w_\Sigma}$. By Lemma \ref{lemma:path-formula}, we obtain
  $\I_\T,w_\Data \models \pathform{w_\Sigma}$. Since every path in
  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
  labeled by final configurations, we obtain that $\I_\T,w_\Data
  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$,
  trivially.  \qed}

In the following, it will be useful to define the set of data
valuations of a word by a formula without predicates. Given a sequence
of input events $\alpha = a_1 \ldots a_n$, let $\substform{\alpha}$ be
the formula obtained from $\pathform{\alpha}$, by successively
replacing in $\accform{\alpha}$: \begin{compactenum}[(a)]
\item\label{it1:substform} each occurrence of a predicate
  $\stamp{q}{i}(t_1,\ldots,t_{\#(q)})$ by the formula
  $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is
  the unique transition rule for $q$ and $a_i$, for all
  $i=0,\ldots,n-1$,
%
\item\label{it2:substform} each occurrence of a predicate
  $\stamp{q}{n}$, such that $q \in Q \setminus F$, by $\top$, and
%
\item\label{it3:substform} each occurrence of a predicate
  $\stamp{q}{n}$, such that $q \in F$, by $\bot$.
\end{compactenum}
Since there are no occurrences of predicates in $\substform{\alpha}$,
for each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
exists an interpretation $\I$ such that $\I,\nu \models
\substform{\alpha}$ if and only if $\J,\nu \models \substform{\alpha}$, 
for every interpretation $\J$. Thus we omit $\I$ and simply write $\nu
\models \substform{\alpha}$ in this case.

\begin{proposition}\label{prop:subst}
  Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
  be the only predicate atom occurring within $\phi$. Then, for every
  valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
  predicate occurrences, the following are
  equivalent: \begin{compactenum}
  \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
      t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
    %
  \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
    \ldots \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow
    \psi$, for some interpretation $\I$.
  \end{compactenum}
  where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
\end{proposition}
\proof{ We give the proof in the case $\phi$ is quantifier-free. The
  generalization to a non-empty quantifier prefix is straightforward.
  ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
  distinguish two cases:
  \begin{compactitem}
  \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
    then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
    $\emptyset$ everywhere else. We have $\I,\nu \models
    q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
    satisfiability under $\I,\nu$, hence $\I,\nu \models
    \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
      t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
    %
  \item else $\nu \not\models \psi[t_1/y_1, \ldots,
    t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
    each predicate symbol with $\emptyset$. Then replacing
    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
    satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
    $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
    y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, trivially.
  \end{compactitem}

>>>>

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. As a shorthand, we write $\stamp{Q}{\leq n}$ (resp.
$\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid q \in Q, i
  \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in [n]}$). For
a formula $\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
variables and state predicates (and only those symbols) are replaced
by their time-stamped counterparts. As a shorthand, we shall write
$q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ (in $\A$) is defined as follows:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\T$}}\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(\vec{y})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
    \text{ occurs on the $n$-th level in $\T$}}$, we have $\I_{\T},
  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
  \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
  execution semantics of $\A$. This is the case because the children
  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
  $(n-1)$-th level of $\T$ form a cube from
  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs on the
  right-hand side. Then it must be the case that, for some node on the
  $(n-1)$-th level of $\G$, labeled with a configuration
  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}

Next, we give a logical characterization of acceptance, relative to a
given sequence of input events $\alpha \in \Sigma^*$. To this end, we
extend the path formula $\pathform{\alpha}$ with a characterization of
the set of final states of $\A$: 
\begin{equation}\label{eq:accform}
\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{n}(\vec{y}) \rightarrow \bot
\end{equation}

\begin{lemma}\label{lemma:acceptance}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
  every word $w \in \Sigma[X]^*$, the following are
  equivalent: \begin{compactenum}
  \item\label{it1:lemma:acceptance} there exists an interpretation
    $\I$ such that $\I,w_\Data \models \accform{w_\Sigma}$, 
    %
  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
  \end{compactenum}
\end{lemma}
\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
  that $\I,w_\Data \models \accform{w_\Sigma}$. By Lemma
  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
  that \begin{inparaenum}[(i)]
    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
      that
    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
      final configurations only. \end{inparaenum} First, assume that
  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
  m < n$. Then there exists a node on the $m$-th level, labeled with
  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
  children. By the definition of the execution semantics of $\A$, we
  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
  for $\nu[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
  $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
  contradicts the fact that $\I=\I_{\T}$ and that
  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
  \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
  similar reasoning as in the above case, we obtain that
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.

  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
  of $\A$ over $w$. We prove that $\I_\T,w_\Data \models
  \accform{w_\Sigma}$. By Lemma \ref{lemma:path-formula}, we obtain
  $\I_\T,w_\Data \models \pathform{w_\Sigma}$. Since every path in
  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
  labeled by final configurations, we obtain that $\I_\T,w_\Data
  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
  trivially.  \qed}

In the following, it will be useful, in several places, to define the
data sequence of an accepted word using the predicate-free formula
defined below:

\begin{definition}\label{def:substform}
Given a sequence of input events $\alpha = a_1 \ldots a_n$, let
$\substform{\alpha}$ be the formula obtained by successively replacing
in $\accform{\alpha}$: \begin{compactenum}[(a)]
\item\label{it1:substform} each occurrence of a predicate
  $\stamp{q}{i}(t_1,\ldots,t_{\#(q)})$ by the formula
  $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
  $q(\vec{y}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is the transition
  rule for $q$ and $a_i$, for all $i=0,\ldots,n-1$,
%
\item\label{it2:substform} each occurrence of a predicate
  $\stamp{q}{n}$, such that $q \in Q \setminus F$, by $\top$, and
%
\item\label{it3:substform} each occurrence of a predicate
  $\stamp{q}{n}$, such that $q \in F$, by $\bot$.
\end{compactenum}
\end{definition}
Since there are no occurrences of predicates in $\substform{\alpha}$,
for each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
exists an interpretation $\I$ such that $\I,\nu \models
\substform{\alpha}$ if and only if $\J,\nu \models
\substform{\alpha}$, for every interpretation $\J$. Thus we omit $\I$
and simply write $\nu \models \substform{\alpha}$ in this case.

\begin{proposition}\label{prop:subst}
  Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
  be the only predicate atom occurring within $\phi$. Then, for every
  valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
  predicate occurrences, the following are
  equivalent: \begin{compactenum}
  \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
      t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
    %
  \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
    \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
    \psi$, for some interpretation $\I$.
  \end{compactenum}
  where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
\end{proposition}
\proof{ We give the proof in the case $\phi$ is quantifier-free. The
  generalization to a non-empty quantifier prefix is straightforward.
  ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
  distinguish two cases:
  \begin{compactitem}
  \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
    then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
    $\emptyset$ everywhere else. We have $\I,\nu \models
    q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
    satisfiability under $\I,\nu$, hence $\I,\nu \models
    \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
      t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
    \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
    %
  \item else $\nu \not\models \psi[t_1/y_1, \ldots,
    t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
    each predicate symbol with $\emptyset$. Then replacing
    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
    satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
    $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
    y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
  \end{compactitem}
